On Devon Allen's Disqualification at the 2022 Track and Field World Championship

### 2023-04-21 Manuscript/manuscript.tex
glmm method

\subsection{GLMM}
Based on an exploratory analysis, the GLMM model that we chose was a gamma
mixed-effect model for the reaction times from the World Championship. The
normal error model on the log-scale was found to be inferior to the gamma model.
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by a
gamma distribution with mean
\[
\mu_{i} = \alpha + z_i,
\]
and variance
\[
\sigma_i
\]
...
\jy{need to add assumptions ...}

The R package \texttt{lme4} allows various linear mixed effects models to be created, and 
was the primary modeling package used for our data analysis \citep{lme4}.
A gamma mixed linear effects model was chosen as the best model to model the data.
Many models were tried and compared to determine what the strongest model.  For
a fixed data set such as the men's pooled data including 2022, the models with
the venue effect included performed better than the models without the venue
effect for each set of comparable models; as indicated by the lower AIC and
higher log likelihood.  Thus, the venue effect is significant and should be
included in the model.  It can also be established that a gamma distribution
better models the data than linear data from comparing the linear model with
venue effect and them gamma model with a venue effect.  The final decision that
was made in model selection was to include link = log in the model.  Despite
doing very little to alter the AIC or log likelihood, the log scale was included
in order to try and better deal with high outliers. %Include Residuals plot?

Once the Gamma mixed effects model was chosen, it needed to be decided which combination
of effects to use.  It was possible to include a venue effect, a batch effect, 
or a year and batch effect.  The batch effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the batch reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the batch
effect, but are all factors that could potentially impact the venue effect.

%Throughout exploratory analysis, gamma provides better AIC
%Spend paragraph on after fitting the model how to figure out 0.1 p value
As the GLMM model has mixed form, the best way to try to gauge the probability
of observing a reaction time is to simulate the distribution of times using a 
large sample.  In order to approximate the distribution, R code was written to 
try and estimate the probabilities of observing a reaction time as extreme as
0.1 seconds.  The code re-paramertizes the gamma distribution to be in terms of
$\mu$ and shape rather than by $\alpha$ and $\beta$.  From the code, it is
possible to calculate the probability of observing a reaction time faster than
a given time.  We want the probability of a reaction time being less than 0.1
seconds in order to gauge if that is a reasonable disqualification barrier.

%T, generate from fitted model explain what
%The R code does.  Simulate from fitted model. Use distribution of model to 
%Figure out the p value of reaction time less than 0.1

@@

%Citations needed
%Include math formulas and the test statistic
%Equation of test statistic is from papers cited by clusrank. Asymptootic dist
%Was found by __ referecne and the implementation is from the clusrank R package

@@

Devon Allen's disqualification at the World Track and Field Championships was
possibly the result of both a faulty timing equipment and Devon Allen
reacting extremely quickly to the start.  The code and data from R showed that
the 2022 World Track and Field Championships were a low outlier compared to
every other year in terms of average reaction time.  Additionally, it was shown
through a gamma mixed effects model that there are significant year and batch
effects that greatly impact reaction time.  However, this practically does not make much sense
as there is no reason for so much random variation without much of a trend.
There has not been a consistent decrease or increase since 1999, much of the data
for reaction times has been random and unpredictable.  Thus while it seems easy
to conclude that Devon Allen was wrongly disqualified, that may not necessarily
be the case.  If there are issues with reaction time such as athletes approaching the 0.10 
second barrier again at the next World Track Championship (August 2023), then World Athletics 
needs to adjust that threshold and allow for faster reaction times so that athletes with superb 
reflexes are not penalized like how Allen was in 2022.  World Athletics did not
do anything about \citet{komi2009iaaf}'s study, but perhaps they should now so
that the story of Devon Allen is not repeated.

>>>>

\subsection{GLMM}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championship, the final model that we chose was a gamma
mixed-effect model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. A random effect for the venue or year was also found to be necessary.
Let $Y_{ij}$ be the reaction time of observation~$j$ in year~$i$.
Conditional on a venue effect $z_i$ of year~i,  $Y_{ij}$ is modeled by 
gamma distribution. In the classic generlized linear model (GLM) sense,
$Y_{ij}$ has conditional mean
\[
g\{E(Y_{ij} | z_i)\} = \alpha + z_i,
\]
and dispersion parameter $\phi$, where $g$ is a known link function and
$z_i$ is a normally distributed random effect with mean zero and
variance~$\sigma^2$. The link function $g$ can be the logarithm function to
ensure the positivity of the conditional mean.
The gamma mixed-effect model can be fit with the \texttt{glmer()} function in R
package \texttt{lme4} \citep{lme4}.

To better understand this model, we can identify the gamma model in terms of the
commonly used shape and scale parameters. For a gamma distribution with
shape~$a$ and scale~$b$, we have mean $\mu = ab$ and variance $v = ab^2$. The
variance as a function of the mean $\mu$ in the GLM sense is
$v(\mu) = \mu^2 / a$, which implies that the dispersion parameter
$\phi = 1 / a$. Therefore, for the gamma mixed-effect model, the conditional
gamma distribution of the reaction time $Y_{ij}$ given the venue effect $z_i$
has shape $a = 1 / \phi$ and scale $b = g^{-1}(\alpha + z_i) \phi$. The marginal
distribution of $Y_{ij}$ is a scale-mixture of gamma distributions, which can be
easily simulated from once the parameters are estimated. A large number of
random numbers generated from the fitted mixture distribution can be used to
approimate the probability of observing a reaction time faster than any given
threshold.  We want the probability of a reaction time being less than 0.1
seconds in order to gauge if that is a reasonable disqualification barrier.

As for the random effect, it was possible to include as well a race effect.
From our exploration, however, the race effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the race reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the race
effect, but are all factors that could potentially impact the venue effect.

@@

@@

Devon Allen's disqualification at the World Track and Field Championships was
possibly the result of both a faulty timing equipment and Devon Allen
reacting extremely quickly to the start.  The code and data from R showed that
the 2022 World Track and Field Championships were a low outlier compared to
every other year in terms of average reaction time.  Additionally, it was shown
through a gamma mixed effects model that there are significant year and race
effects that greatly impact reaction time.  However, this practically does not make much sense
as there is no reason for so much random variation without much of a trend.
There has not been a consistent decrease or increase since 1999, much of the data
for reaction times has been random and unpredictable.  Thus while it seems easy
to conclude that Devon Allen was wrongly disqualified, that may not necessarily
be the case.  If there are issues with reaction time such as athletes approaching the 0.10 
second barrier again at the next World Track Championship (August 2023), then World Athletics 
needs to adjust that threshold and allow for faster reaction times so that athletes with superb 
reflexes are not penalized like how Allen was in 2022.  World Athletics did not
do anything about \citet{komi2009iaaf}'s study, but perhaps they should now so
that the story of Devon Allen is not repeated.
